On Optimality of Finite Element Integration

### 2015-12-10 TOMS_Paper/assembly.tex
Add examples to sharing elimination. Minor fixes

The need for rapid implementation of high performance, robust, and portable finite element methods has led to approaches based on automated code generation. This has been proven successful in the context of the FEniCS (\cite{Fenics}) and Firedrake (\cite{firedrake-code}) projects. In these frameworks, the weak variational form of a problem is expressed at high level by means of a domain-specific language. The mathematical specification is manipulated by a form compiler that generates a representation of assembly operators. By applying these operators to an element in the discretized domain, a local matrix and a local vector, which represent the contributions of that element to the equation approximated solution, are computed. The code for assembly operators must be carefully optimized: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, or polynomial order of the chosen function spaces, the operation count increases, with the result that assembly often accounts for a significant fraction of the overall runtime.

@@

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 7pt \subfigure[With spatial sharing]{\label{code:multi_loopnest_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}}}
~~~~~
{\subfigcapskip = 7pt \subfigure[With temporal sharing]{\label{code:multi_loopnest_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_int.code}}}
~~~~~
{\subfigcapskip = 2pt \subfigure[Optimal form]{\label{code:multi_loopnest_c}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}}}
\end{CenteredBox}\caption{Reducing a simple multilinear loop nest to optimal form.}\label{code:multi_loopnest}\end{figure}

@@

Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} discuss the structural (i.e., inherent in finite element integration) and algebraic (i.e., related to the use of differential and vector operators) properties that can characterize our expressions. Section~\ref{sec:se-algo} presents the sharing elimination algorithm. We recall that we are assuming a generic finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.

@@

\begin{Algo}[Sharing elimination]
\label{algo:sharing-elimination}
\normalfont
We introduce the following notation and terminology
\begin{itemize}
\item $|.|$: a generic ``sizeof'' operator, denoting the cardinality of a set or the extent of a loop
\item $c_o$: the cost (i.e., number of operations) of an expression $o$
\item A \textit{multilinear operand} is any $\alpha_{eij}$ or $\beta_{eik}$ in the input expression; a \textit{multilinear symbol} is any symbol appearing in a multilinear operand (e.g., test functions)
\end{itemize}

The initialization step consists of a depth-first visit of the expression tree with recursive collection of identical sub-expressions. This ensures the uniqueness of operands. For example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$.

We perform a second depth-first visit of the expression tree to collect all multilinear operands and partition them into $p$ disjoint sets -- that is $\mathbb{P} = \lbrace P_0, P_1, ..., P_{p-1}\rbrace$ such that $\cup_{i=1}^p P_i = \mathbb{P}$ and $\forall i, j = 0,..., p-1$, $i \neq j$, $P_i \cap P_j = \emptyset$. The multilinear operands in a partition $P_i$ share a set of symbols (e.g., all operands in which first order derivatives of test functions appear), whereas there is no sharing across different partitions. We have to establish:
\begin{enumerate}
\item an ordering for $\mathbb{P}$, such that if $P_i \prec P_j$, $P_i$ is transformed before $P_j$, and
\item a scheduling strategy for each $P_i$
\end{enumerate}

We start with point 1). We further partition $\mathbb{P}$ into two disjoint sets $\mathbb{P}^{j}$ and $\mathbb{P}^{k}$, which logically separates partitions based on loop dependency. We then order $\mathbb{P}$ lexicographically, with the position of $P^{l}$ given by ${<}|\mathbb{P}^{l}|,\ |L_l| \sum_{o \in P^{l}} c_o{>}$. Intuitively, this simply means that operands appearing more often in the expression will be factorized first, which is essential to minimize the operation count within a loop.

We now tackle point 2). For each partition, we have to decide between two options, as already explained in Section~\ref{sec:se-rln}. We recall them briefly: (i) applying generalized code motion, SCSE (if possible), and CCSE (if necessary); (ii) expanding products to expose spatial sharing, which enables factorization and, in turn, generalized code motion. For this choice, we adopt the following criteria:
\begin{enumerate}
\item Let $S_P$ be set of multilinear symbols in $P$. If $|P| < |S_P|$, (i) is applied; if $|P| > |S_P|$, (ii) is applied; if $|P| = |S_P|$, we rely on criterion 2), which attempts to optimize for the outer loops
\item Let $n_{o}$ and $n_{s}$ denote, respectively, the number of times a multilinear operand $o$ and a multilinear symbol $s$ appear in the input expression. Let $X$ be the set of multilinear operands, and let $Y$ be the set of multilinear symbols. Assume, for this example, that $P$ depends on $[L_i, L_k]$ (which we know, from Figure~\ref{code:loopnest}, to have extent $M$ and $O$, respectively). We observe that the increase in operation count at the level of the outer loops would be: $M (O \sum_{o \in X} c_{o} + \sum_{o \in X} (n_{o} - 1))$ if (i) were applied; $M \sum_{s \in Y} (n_{s} - 1)$ if (ii) were applied. We then take (i) or (ii) depending on which quantity is minimum.
\end{enumerate}

%TODO need intuition here!

It remains to schedule $L_i$-dependent expressions; that is, operations that, logically, do not belong to the multilinear loop nest. We perform a last depth-first visit of the expression tree and, for each of such expressions, we apply sharing elimination following a strategy which is very close to point (2) above. In doing this, we make the assumption that an $L_i$-dependent expression is the result of a structured tensor operation. If this were not the case, the factorization is chosen greedily (i.e., eventually, some code motion opportunities will be exposed, but this may lead to a local optimum).
\end{Algo}

@@

We start with an example. Consider again the bilinear form implementation in Figure~\ref{code:loopnest}. We pose the following question: are we able to identify sub-expressions within $F$ for which the reduction induced by $L_i$ can be pre-evaluated, thus obtaining a decrease in operation count proportional to the size of $L_i$, $M$? The transformation we look for is exemplified in Figures~\ref{code:loopnest_red} and~\ref{code:loopnest_nored}; Figure~\ref{code:loopnest_red}, which represents the input, can be seen as a simple instance of the abstract loop nest in Figure~\ref{code:loopnest}.

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 19pt \subfigure[With reduction]{\label{code:loopnest_red}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_red.code}}}
~~~~~~~~~~
{\subfigcapskip = 2pt \subfigure[Pre-evaluated reduction]{\label{code:loopnest_nored}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_nored.code}}}
\end{CenteredBox}\caption{Exposing (through factorization) and pre-evaluating a reduction.}\label{code:loopnest_rednored}\end{figure}

@@

The pre-evaluation of a monomial introduces some critical issues:
\begin{enumerate}
\item In contrast to what happens with hoisting in multilinear loop nests, the temporary variable size is proportional to the number and trip counts of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $N O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L N O$ for those depending on $[L_e, L_i, L_j, L_k] $). This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual execution time.
\item The transformations exposing $[L_i, L_j, L_k]$-dependent terms increase, in general, the arithmetic complexity (e.g., expansion may increase the operation count). This could outweigh the gain due to pre-evaluation.
\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities: sharing elimination inhibits pre-evaluation, whereas pre-evaluation generally exposes further sharing elimination opportunities.
\end{enumerate}

@@

\begin{itemize}
\item The fact that $L \gg M, N, O$ suggests we should be cautious about hoisting mesh-dependent (i.e., $L_e$-dependent) expressions. Imagine $\Lambda$ is enclosed in a time stepping loop. One could think of exposing (through some transformations) and hoisting any time-invariant sub-expressions to minimize redundant computation at every time step. The working set size could then increase by a factor $L$. The gain in number of operations executed could therefore be outweighed, from a runtime viewpoint, by a much larger memory pressure.
\item For certain forms and discretizations, aggressive hoisting can make the working set exceed the size of some level of local memory (e.g. the last level of private cache on a conventional CPU, the shared memory on a GPU). For example, as explained in Section~\ref{sec:pre-evaluation}, pre-evaluating geometry-independent expressions outside of $\Lambda$ requires temporary arrays of size $N O$ for bilinear forms and of size $N$ (or $O$) for linear forms. This can sometimes break such a ``local memory threshold''. In our experiments (Section~\ref{sec:perf-results-forms}) we will carefully study this aspect.
\end{itemize}

>>>>

The need for rapid implementation of high performance, robust, and portable finite element methods has led to approaches based on automated code generation. This has been proven successful in the context of the FEniCS (\cite{Fenics}) and Firedrake (\cite{firedrake-paper}) projects. In these frameworks, the weak variational form of a problem is expressed at high level by means of a domain-specific language. The mathematical specification is manipulated by a form compiler that generates a representation of assembly operators. By applying these operators to an element in the discretized domain, a local matrix and a local vector, which represent the contributions of that element to the equation approximated solution, are computed. The code for assembly operators must be carefully optimized: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, or polynomial order of the chosen function spaces, the operation count increases, with the result that assembly often accounts for a significant fraction of the overall runtime.

@@

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 11pt \subfigure[With spatial sharing]{\label{code:multi_loopnest_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}}}
~~~~~
{\subfigcapskip = 11pt \subfigure[With temporal sharing]{\label{code:multi_loopnest_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_int.code}}}
~~~~~
{\subfigcapskip = 2pt \subfigure[Optimal form]{\label{code:multi_loopnest_c}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}}}
\end{CenteredBox}\caption{Reducing a simple multilinear loop nest to optimal form.}\label{code:multi_loopnest}\end{figure}

@@

Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} discuss the structural (i.e., inherent in finite element integration) and algebraic (i.e., related to the use of differential and vector operators) properties that characterize our expressions. Section~\ref{sec:se-algo} presents the sharing elimination algorithm. We recall that we are assuming a generic finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.

@@

\begin{Algo}[Sharing elimination]
\label{algo:sharing-elimination}
\normalfont
We introduce the following notation and terminology
\begin{itemize}
\item $|.|$: a generic ``sizeof'' operator, denoting the cardinality of a set or the extent of a loop
\item $c_o$: the cost (i.e., number of operations) of an expression $o$
\item A \textit{multilinear operand} is any $\alpha_{eij}$ or $\beta_{eik}$ in the expression; a \textit{multilinear symbol} is a symbol appearing in a multilinear operand depending on $L_j$ or $L_k$ (e.g., test functions, trial functions, first order derivatives of test functions, etc.)
\end{itemize}

The initialization step consists of a depth-first visit of the expression tree with recursive collection of identical sub-expressions. For example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$.

We perform a second depth-first visit of the expression tree to collect all multilinear operands and partition them into $p$ disjoint sets -- that is $\mathbb{P} = \lbrace P_0, P_1, ..., P_{p-1}\rbrace$ such that $\cup_{i=1}^p P_i = \mathbb{P}$ and $\forall i, j = 0,..., p-1$, $i \neq j$, $P_i \cap P_j = \emptyset$. The multilinear operands in a partition $P_i$ share a set of multilinear symbols, whereas there is no sharing across different partitions. We have to establish:
\begin{enumerate}
\item an ordering for $\mathbb{P}$, such that if $P_i \prec P_j$, then $P_i$ is scheduled before $P_j$, and
\item a scheduling strategy for each $P_i$
\end{enumerate}

We start with point 1). We further partition $\mathbb{P}$ into two disjoint sets $\mathbb{P}^{j}$ and $\mathbb{P}^{k}$, which logically separates partitions based on loop dependency. We then order $\mathbb{P}$ lexicographically, with the position of $P \in \mathbb{P}^j$ given by ${<}|\mathbb{P}^{j}|,\ |L_j| \sum_{o \in P} c_o{>}$ (and analogously for $P \in \mathbb{P}^k$). Intuitively, this simply means that operands appearing more often in the expression are handled first. We will later elaborate on this.

We now tackle point 2). For each partition, we have two options, as already explained in Section~\ref{sec:se-rln}. We recall them briefly: (i) applying generalized code motion, SCSE (if possible), and CCSE; (ii) expanding products to expose spatial sharing, which enables factorization and, in turn, generalized code motion. We adopt the following criteria:
\begin{enumerate}
\item Let $S_P$ be the set of multilinear symbols in $P$. If $|P| < |S_P|$, (i) is applied; if $|P| > |S_P|$, (ii) is applied; if $|P| = |S_P|$, we rely on criterion 2), which attempts to optimize for the outer loops.
\item Let $n_{o}$ and $n_{s}$ denote, respectively, the number of times a multilinear operand $o$ and a multilinear symbol $s$ appear in the input expression. Assume that $P$ depends on $[L_i, L_k]$. The increase in operation count at the level of the outer loops is: if (i) were applied, $|L_i| (|L_k| \sum_{o \in P} c_{o} + \sum_{o \in P} (n_{o} - 1))$; if (ii) were applied, $|L_i| \sum_{s \in P_S} (n_{s} - 1)$. We pick (i) or (ii) such that the increase is minimum. Intuitively, this is a way of estimating how two different factorizations affect the operation count, given the fact (criterion 1) that they both minimize the cost of the innermost loop.
\end{enumerate}

It remains to schedule $L_i$-dependent expressions. We perform a last depth-first visit of the expression tree and, for each of such expressions, we apply sharing elimination following a strategy which is very close to point (2) above. In doing this, we speculatively assume that all expressions result from structured tensor operations. If this were not the case, factorization would be applied greedily (i.e., eventually, some code motion opportunities will be exposed, but this may lead to a local optimum).
\end{Algo}

\subsubsection{Examples}
Figure~\ref{code:poisson} applies Algorithm~\ref{algo:sharing-elimination} to the bilinear form arising from a Poisson equation in 2D. We observe that $P_0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P_1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$. Clearly, $\mathbb{P} = \lbrace P_0, P_1 \rbrace$, with $\mathbb{P}^k = \lbrace P_0 \rbrace$ and $\mathbb{P}^j = \lbrace P_1 \rbrace$. A much more complex example, using the hyperelastic model evaluated in Section~\ref{sec:perf-results}, is made available\footnote{Sharing elimination example - application to a hyperelastic model: \url{https://gist.github.com/FabioLuporini/14e79457d6b15823c1cd}}.
\begin{figure}\begin{CenteredBox}
{\subfigcapskip = 7pt \subfigure[Normal form]{\label{code:poisson_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_pre_se.code}}}
~~~~~~~~~~
{\subfigcapskip = 7pt \subfigure[Optimal form]{\label{code:poisson_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/poisson_after_se.code}}}
\end{CenteredBox}\caption{Applying sharing elimination to the bilinear form $J^{-T} \nabla u \cdot J^{-T} \nabla v$ arising from a Poisson equation in 2D.}\label{code:poisson}\end{figure}

@@

We start with an example. Consider again the bilinear form implementation in Figure~\ref{code:loopnest}. We pose the following question: are we able to identify sub-expressions within $F$ for which the reduction induced by $L_i$ can be pre-evaluated, thus obtaining a decrease in operation count proportional to the size of $L_i$, $I$? The transformation we look for is exemplified in Figures~\ref{code:loopnest_red} and~\ref{code:loopnest_nored}; Figure~\ref{code:loopnest_red}, which represents the input, can be seen as a simple instance of the abstract loop nest in Figure~\ref{code:loopnest}.

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 19pt \subfigure[With reduction]{\label{code:loopnest_red}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_red.code}}}
~~~~~~~~~~
{\subfigcapskip = 5pt \subfigure[Pre-evaluated reduction]{\label{code:loopnest_nored}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_nored.code}}}
\end{CenteredBox}\caption{Exposing (through factorization) and pre-evaluating a reduction.}\label{code:loopnest_rednored}\end{figure}

@@

The pre-evaluation of a monomial introduces some critical issues:
\begin{enumerate}
\item In contrast to what happens with hoisting in multilinear loop nests, the temporary variable size is proportional to the number and trip counts of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $J K$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $E J K$ for those depending on $[L_e, L_i, L_j, L_k] $). This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual execution time.
\item The transformations exposing $[L_i, L_j, L_k]$-dependent terms increase, in general, the arithmetic complexity (e.g., expansion may increase the operation count). This could outweigh the gain due to pre-evaluation.
\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities: sharing elimination inhibits pre-evaluation, whereas pre-evaluation generally exposes further sharing elimination opportunities.
\end{enumerate}

@@

\begin{itemize}
\item The fact that $E \gg I, J, K$ suggests we should be cautious about hoisting mesh-dependent (i.e., $L_e$-dependent) expressions. Imagine $\Lambda$ is enclosed in a time stepping loop. One could think of exposing (through some transformations) and hoisting any time-invariant sub-expressions to minimize redundant computation at every time step. The working set size could then increase by a factor $E$. The gain in number of operations executed could therefore be outweighed, from a runtime viewpoint, by a much larger memory pressure.
\item For certain forms and discretizations, aggressive hoisting can make the working set exceed the size of some level of local memory (e.g. the last level of private cache on a conventional CPU, the shared memory on a GPU). For example, as explained in Section~\ref{sec:pre-evaluation}, pre-evaluating geometry-independent expressions outside of $\Lambda$ requires temporary arrays of size $J K$ for bilinear forms and of size $J$ (or $K$) for linear forms. This can sometimes break such a ``local memory threshold''. In our experiments (Section~\ref{sec:perf-results-forms}) we will carefully study this aspect.
\end{itemize}
